Flattering Advice: Avoiding Disappointment in Advice-Giving

### 2025-01-28 manuscript/manuscript.qmd
minor changes

Next, we recruited `r scales::comma(dim(study1_time1)[1])` participants
for the role of Advisees. They completed the same 10-item quiz, also
earning five cents for each correct answer. We then asked them to
express a non-binding preference for whether they would like to compete
against the High or Low Performer Group. We informed them that they
would be invited back at a later date when they could make a binding
decision and they could earn a bonus based on whether their score on
this quiz was equal to or higher than a randomly selected member from
their comparison group. If they picked the High Performer Group and had
an equal or higher score than a randomly selected member of that group,
they would earn a bonus of 50 cents. If they picked the Low Performer
Group and outperformed a randomly selected member of that group, they
could earn a bonus of 20 cents. If their score was lower in this
comparison, they would not earn an additional bonus. We randomized which
of the two groups (High or Low Performers) was selected by default, and
Advisees were free to select the other group. They also made a guess as
to their score, and the survey concluded with basic demographic
questions.

@@

Finally, we invited Advisees back for the follow-up survey. Following
our preregistration, we kept the survey open for 7 days. In total,
`r scales::comma(dim(study1_time3)[1])` Advisees returned. The short
survey reminded them of the task they completed in the previous survey,
informed them that other participants from Prolific had observed their
real score and given them advice on which group to compete against, and
finally reminded them of how many questions they guessed they had
answered correctly. They were not informed of their true score or the
score of the groups they could compete against. Participants then
observed the advice from a randomly selected adviser and made their
decision.

@@

study1_time2_advice %>%
  mutate(Treatment=factor(Treatment, levels=c('High','Low'), labels = c('High default', 'Low default'))) %>%
  group_by(Condition, Treatment) %>%
  summarize(avg_count = mean(Advice), ci = qnorm(0.975)*sd(Advice)/sqrt(dplyr::n())) %>%
  ggplot2::ggplot(ggplot2::aes(x = Condition, y = avg_count, fill = Treatment)) +
  ggplot2::geom_col(position = 'dodge') +
  ggplot2::geom_errorbar(ggplot2::aes(ymin = avg_count - ci, ymax = avg_count + ci), 
                         position = ggplot2::position_dodge(0.9), linewidth = 0.2, width = 0.1) +
  # ggtitle("Advice: Meeting Expectations vs. Absent Expectations") +
  scale_fill_manual(values = c("#2A629A",'#FF7F3E')) +
  scale_y_continuous("P(Compete Against High Performer)",
                     labels = scales::percent) +
  geom_text(aes(x = Condition, y = 0.02,
                label = paste0(round(avg_count*100), "%")),
            position = position_dodge(0.9),
            color = "white") +
  papaja::theme_apa() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "grey"),
        axis.text.x = element_text(size = 12),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none", # No legend
        plot.title = element_text(hjust = 0.5, size = 16)) +
  scale_x_discrete(labels = c("Expectation Shown", "Expectation Hidden")) +
   geom_text(aes(label=Treatment, y = -0.01), position = position_dodge(0.90))

@@

The study takes place in three stages. First, a group of advisees upload
photos of themselves ("selfies") and are grouped with nine other
participants of the same sex We then recruit participants of the
opposing sex to rank them from most to least attractive and to provide
advice. Specifically, they advise the participant they ranked as the 7th
most attractive (i.e., 4th least attractive) on what rank they should
bet they were ranked by a larger group of raters. Advisers were randomly
assigned to two treatments, receiving a bonus payment either if the
advisee guessed their rank accurately or if the advisee evaluated the
adviser as likeable as measured by a scale response. We hypothesize that
advisers incentivized to be liked will recommend betting on a lower
rank, i.e. that the advisee is more attractive.

@@

Next, we recruit `r scales::comma(dim(study2_time2)[1])` participants
from Prolific for the role of advisers ($M_{\text{Age}}$ =
`r mean(study2_time2$age)` years;
`r round(mean(study2_time2$gender == "Female")*100, 2)`% Female). We
first collect demographic information, then match them to a group of the
opposite sex. They rank selfies from ten participants from most to least
attractive using a drop-down menu next to each picture. Because of a
limitation with the survey software, we could not validate that each
rank is given only once, and we remove 115 participants who failed to
follow instructions and did not provide a complete ranking.[^2]

@@

Finally, we examine whether flattering advice indeed leads to more
positive evaluations of advisers, or whether flattering advice is
dismissed as insencere. Following our preregistration, we average the
ratings on likeability, warmth, friendliness, and good-naturedness to
create a scale of likeability ($\alpha$ =
`r cronbach.alpha(study2_time3[, 5:8])$alpha`); and we create a scale of
trustworthiness by averaging the ratings of trustworthiness and
sincerity ($\alpha$ = `r cronbach.alpha(study2_time3[, 9:10])$alpha`).

>>>>

Next, we recruited `r scales::comma(dim(study1_time1)[1])` participants
for the role of Advisees. They completed the same 10-item quiz, also
earning five cents for each correct answer. We then asked them to
express a non-binding preference for whether they would like to compete
against the High or Low Performer Group. We informed them that they
would be invited back at a later date when they could make a binding
decision and they could earn a bonus based on whether their score on
this quiz was equal to or higher than a randomly selected member from
their comparison group. If they picked the High Performer Group and had
an equal or higher score than a randomly selected member of that group,
they would earn a bonus of 50 cents. If they picked the Low Performer
Group and outperformed a randomly selected member of that group, they
could earn a bonus of 20 cents. If their score was lower in this
comparison, they would not earn an additional bonus. We randomized which
of the two groups (High or Low Performers) was selected by default, and
Advisees were free to select the other group. The survey concluded 
with basic demographic questions.

@@

Finally, we invited Advisees back for the follow-up survey. Following
our preregistration, we kept the survey open for 7 days. In total,
`r scales::comma(dim(study1_time3)[1])` Advisees returned. The short
survey reminded them of the task they completed in the previous survey,
informed them that other participants from Prolific had observed their
real score and given them advice on which group to compete against, and, 
finally, reminded them of their initial choice of comparison group. 
They were not informed of their true score or the
score of the groups they could compete against. Participants then
observed the advice from a randomly selected adviser and made their
decision.

@@

study1_time2_advice %>%
  mutate(Treatment=factor(Treatment, levels=c('High','Low'), labels = c('High default', 'Low default'))) %>%
  group_by(Condition, Treatment) %>%
  summarize(avg_count = mean(Advice), ci = qnorm(0.975)*sd(Advice)/sqrt(dplyr::n())) %>%
  ggplot2::ggplot(ggplot2::aes(x = Condition, y = avg_count, fill = Treatment)) +
  ggplot2::geom_col(position = 'dodge') +
  ggplot2::geom_errorbar(ggplot2::aes(ymin = avg_count - ci, ymax = avg_count + ci), 
                         position = ggplot2::position_dodge(0.9), linewidth = 0.2, width = 0.1) +
  # ggtitle("Advice: Meeting Expectations vs. Absent Expectations") +
  scale_fill_manual(values = c("#2A629A",'#FF7F3E')) +
  scale_y_continuous("P(Compete Against High Performer)",
                     labels = scales::percent) +
  geom_text(aes(x = Condition, y = 0.02,
                label = paste0(round(avg_count*100), "%")),
            position = position_dodge(0.9),
            color = "white") +
  papaja::theme_apa() +
  theme(panel.grid = element_blank(),
        panel.border = element_blank(),
        axis.line = element_line(colour = "grey"),
        axis.text.x = element_text(size = 12, face = "bold",color = "black"),
        axis.title.y = element_text(size = 12, face = "bold"),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none", # No legend
        plot.title = element_text(hjust = 0.5, size = 16)) +
  scale_x_discrete(labels = c("Expectation Shown", "Expectation Hidden")) +
   geom_text(aes(label=Treatment, y = -0.02), position = position_dodge(0.90), color = "#464747"  )

@@

The study takes place in three stages. First, a group of advisees upload
photos of themselves ("selfies") and are grouped with nine other
participants of the same sex. We then recruit participants of the
opposing sex to rank them from most to least attractive and to provide
advice. Specifically, they advise the participant they ranked as the 7th
most attractive (i.e., 4th least attractive) on what rank they should
bet they were ranked by a larger group of raters. Advisers were randomly
assigned to two treatments, receiving a bonus payment either if the
advisee guessed their rank accurately or if the advisee evaluated the
adviser as likeable as measured by a scale response. We hypothesize that
advisers incentivized to be liked will recommend betting on a lower
rank, i.e. that the advisee is more attractive.

@@

Next, we recruit `r scales::comma(dim(study2_time2)[1])` participants
from Prolific for the role of advisers ($M_{\text{Age}}$ =
`r round(mean(study2_time2$age),2)` years;
`r round(mean(study2_time2$gender == "Female")*100, 2)`% Female). We
first collect demographic information, then match them to a group of the
opposite sex. They rank selfies of ten participants from most to least 
attractive by entering their rankings into a text box next to each picture. 
Because of a limitation with the survey software, we could not validate that each
rank is given only once, and we remove 115 participants who failed to
follow instructions and did not provide a complete ranking.[^2]

@@

Finally, we examine whether flattering advice indeed leads to more
positive evaluations of advisers, or whether flattering advice is
dismissed as insencere. Following our preregistration, we average the
ratings on likeability, warmth, friendliness, and good-naturedness to
create a scale of likeability ($\alpha$ =
`r round(cronbach.alpha(study2_time3[, 5:8])$alpha,2)`); and we create a scale of
trustworthiness by averaging the ratings of trustworthiness and
sincerity ($\alpha$ = `r round(cronbach.alpha(study2_time3[, 9:10])$alpha,2)`).

### 2025-01-28 manuscript/manuscript.tex
minor changes

People often go to great lengths to avoid unpleasant information,
deriving belief-based utility from maintaining favorable views of
themselves (Golman et al., 2017; Ho et al., 2021; Loewenstein \& Molnar,
2018). Moreover, they often prefer advisers who withhold bad news from
them (Shalvi et al., 2019) and punish those who communicate bad news
(John et al., 2019). Advisers, anticipating these preferences and
consequences face a dilemma: Should they provide the accurate but
potentially painful truth, or shade their recommendations to preserve
the advisee's mood and the adviser's own standing? Advisers who fear
they will be blamed or disliked for undermining someone's confidence may
reasonably avoid candid feedback, particularly if they don't incur a
cost when the advisee makes a mistake.

In our experiments, however, advisees have no opportunity to punish the
adviser, and advisers' financial incentives are linked to the decisions
the advisees make. We propose that advisers nonetheless have reason to
provide flattering advise because they recognize the psychological toll
of disappointing news. As a result, they may similarly incur a hedonic
cost for delivering unpleasant information that they anticipate may
distress the advisee. Thus, even in the anonymized context of an online
experiment, we hypothesize (and find) that advisers are relucant to
disappoint advisees and advice is thus biased upward, away from accuracy
and toward confirming advisees' priors. We show that this reduces the
quality of advice---but that advisees nonetheless view flattering
advisers more favorably.

@@

Next, we recruited 201 participants for the role of Advisees. They
completed the same 10-item quiz, also earning five cents for each
correct answer. We then asked them to express a non-binding preference
for whether they would like to compete against the High or Low Performer
Group. We informed them that they would be invited back at a later date
when they could make a binding decision and they could earn a bonus
based on whether their score on this quiz was equal to or higher than a
randomly selected member from their comparison group. If they picked the
High Performer Group and had an equal or higher score than a randomly
selected member of that group, they would earn a bonus of 50 cents. If
they picked the Low Performer Group and outperformed a randomly selected
member of that group, they could earn a bonus of 20 cents. If their
score was lower in this comparison, they would not earn an additional
bonus. We randomized which of the two groups (High or Low Performers)
was selected by default, and Advisees were free to select the other
group. They also made a guess as to their score, and the survey
concluded with basic demographic questions.

@@

Finally, we invited Advisees back for the follow-up survey. Following
our preregistration, we kept the survey open for 7 days. In total, 176
Advisees returned. The short survey reminded them of the task they
completed in the previous survey, informed them that other participants
from Prolific had observed their real score and given them advice on
which group to compete against, and finally reminded them of how many
questions they guessed they had answered correctly. They were not
informed of their true score or the score of the groups they could
compete against. Participants then observed the advice from a randomly
selected adviser and made their decision.

@@

The study takes place in three stages. First, a group of advisees upload
photos of themselves (``selfies'') and are grouped with nine other
participants of the same sex We then recruit participants of the
opposing sex to rank them from most to least attractive and to provide
advice. Specifically, they advise the participant they ranked as the 7th
most attractive (i.e., 4th least attractive) on what rank they should
bet they were ranked by a larger group of raters. Advisers were randomly
assigned to two treatments, receiving a bonus payment either if the
advisee guessed their rank accurately or if the advisee evaluated the
adviser as likeable as measured by a scale response. We hypothesize that
advisers incentivized to be liked will recommend betting on a lower
rank, i.e.~that the advisee is more attractive.

@@

Next, we recruit 472 participants from Prolific for the role of advisers
(\(M_{\text{Age}}\) = 41.0275424 years; 49.79\% Female). We first
collect demographic information, then match them to a group of the
opposite sex. They rank selfies from ten participants from most to least
attractive using a drop-down menu next to each picture. Because of a
limitation with the survey software, we could not validate that each
rank is given only once, and we remove 115 participants who failed to
follow instructions and did not provide a complete ranking.\footnote{Moreover,
  we could not collect data from any participant who did not select any
  participant as the 7th most attractive participant, as that image
  would be shown on subsequent pages.}

@@

Finally, we examine whether flattering advice indeed leads to more
positive evaluations of advisers, or whether flattering advice is
dismissed as insencere. Following our preregistration, we average the
ratings on likeability, warmth, friendliness, and good-naturedness to
create a scale of likeability (\(\alpha\) = 0.9243334); and we create a
scale of trustworthiness by averaging the ratings of trustworthiness and
sincerity (\(\alpha\) = 0.8740347).

>>>>

People go to great lengths to avoid unpleasant information, People often
go to great lengths to avoid unpleasant information, deriving
belief-based utility from maintaining favorable views of themselves
(Golman et al., 2017; Ho et al., 2021; Loewenstein \& Molnar, 2018).
Moreover, they often prefer advisers who withhold bad news from them
(Shalvi et al., 2019) and punish those who communicate bad news (John et
al., 2019). Advisers, anticipating these preferences and consequences
face a dilemma: Should they provide the accurate but potentially painful
truth, or shade their recommendations to preserve the advisee's mood and
the adviser's own standing? Advisers who fear they will be blamed or
disliked for undermining someone's confidence may reasonably avoid
candid feedback, particularly if they don't incur a cost when the
advisee makes a mistake.

In our experiments, however, advisees have no opportunity to punish the
adviser, and advisers' financial incentives are linked to the decisions
the advisees make. We propose that advisers nonetheless have reason to
provide flattering advise because they recognize the psychological toll
of disappointing news. As a result, they may similarly incur a hedonic
cost for delivering unpleasant information that they anticipate may
distress the advisee. Thus, even in the anonymized context of an online
experiment, we hypothesize (and find) that advisers are relucant to
disappoint advisees and advice is thus biased upward, away from
accuracy. We show that this reduces the disappoint advisees and advice
is thus biased upward, away from accuracy and toward confirming
advisees' priors. We show that this reduces the quality of advice---but
that advisees nonetheless view flattering advisers more favorably.

@@

Next, we recruited 201 participants for the role of Advisees. They
completed the same 10-item quiz, also earning five cents for each
correct answer. We then asked them to express a non-binding preference
for whether they would like to compete against the High or Low Performer
Group. We informed them that they would be invited back at a later date
when they could make a binding decision and they could earn a bonus
based on whether their score on this quiz was equal to or higher than a
randomly selected member from their comparison group. If they picked the
High Performer Group and had an equal or higher score than a randomly
selected member of that group, they would earn a bonus of 50 cents. If
they picked the Low Performer Group and outperformed a randomly selected
member of that group, they could earn a bonus of 20 cents. If their
score was lower in this comparison, they would not earn an additional
bonus. We randomized which of the two groups (High or Low Performers)
was selected by default, and Advisees were free to select the other
group. The survey concluded with basic demographic questions.

@@

Finally, we invited Advisees back for the follow-up survey. Following
our preregistration, we kept the survey open for 7 days. In total, 176
Advisees returned. The short survey reminded them of the task they
completed in the previous survey, informed them that other participants
from Prolific had observed their real score and given them advice on
which group to compete against, and, finally, reminded them of their
initial choice of comparison group. They were not informed of their true
score or the score of the groups they could compete against.
Participants then observed the advice from a randomly selected adviser
and made their decision.

@@

The study takes place in three stages. First, a group of advisees upload
photos of themselves (``selfies'') and are grouped with nine other
participants of the same sex. We then recruit participants of the
opposing sex to rank them from most to least attractive and to provide
advice. Specifically, they advise the participant they ranked as the 7th
most attractive (i.e., 4th least attractive) on what rank they should
bet they were ranked by a larger group of raters. Advisers were randomly
assigned to two treatments, receiving a bonus payment either if the
advisee guessed their rank accurately or if the advisee evaluated the
adviser as likeable as measured by a scale response. We hypothesize that
advisers incentivized to be liked will recommend betting on a lower
rank, i.e.~that the advisee is more attractive.

@@

Next, we recruit 472 participants from Prolific for the role of advisers
(\(M_{\text{Age}}\) = 41.03 years; 49.79\% Female). We first collect
demographic information, then match them to a group of the opposite sex.
They rank selfies of ten participants from most to least attractive by
entering their rankings into a text box next to each picture. Because of
a limitation with the survey software, we could not validate that each
rank is given only once, and we remove 115 participants who failed to
follow instructions and did not provide a complete ranking.\footnote{Moreover,
  we could not collect data from any participant who did not select any
  participant as the 7th most attractive participant, as that image
  would be shown on subsequent pages.}

@@

Finally, we examine whether flattering advice indeed leads to more
positive evaluations of advisers, or whether flattering advice is
dismissed as insencere. Following our preregistration, we average the
ratings on likeability, warmth, friendliness, and good-naturedness to
create a scale of likeability (\(\alpha\) = 0.92); and we create a scale
of trustworthiness by averaging the ratings of trustworthiness and
sincerity (\(\alpha\) = 0.87).
